---
title: "Group Project 1: Group 2"
author: "Alyssa Bautista, Raquel Gonzalez, Celia Lopez, and Quynh Pham"
date: "2026-03-08"
output: html_document
---

## Data Management & Wrangling

### Data Import

```{r}
# Import the dataset using a relative path
hints_data <- read.csv("data/HINTSData_2020_clean.csv")
# Preview the data to ensure it loaded correctly
head(hints_data)
```

### Data Cleaning

```{r}
# Check for missing values (NAs) by nesting the sum() and is.na() functions
sum(is.na(hints_data))

# Factor the categorical variables
# Extract and replace the Quality Care data and use the factor() function to assign levels and labels to the data
hints_data$QualityCare <- factor(hints_data$QualityCare, levels = c(1, 2, 3, 4, 5), labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"), ordered = TRUE)

# Extract and replace the BirthGender data and use the factor() function to assign levels and labels to the data
hints_data$BirthGender <- factor(hints_data$BirthGender, levels = c(1, 2), labels = c("Male", "Female"), ordered = TRUE)

# Extract and replace the smokeStat data and use the factor() function to assign levels and labels to the data
hints_data$smokeStat <- factor(hints_data$smokeStat, levels = c(1, 2, 3), labels = c("Current", "Former", "Never"), ordered = TRUE)

# Extract and replace the RaceEthn5 data and use the factor() function to assign levels and labels to the data
hints_data$RaceEthn5 <- factor(hints_data$RaceEthn5, levels = c(1, 2, 3, 4, 5), labels = c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic", "Non-Hispanic Asian", "Non-Hispanic Other"), ordered = TRUE)

```

### Extracting Relevant Variables

```{r}
# Use the subset() function to create a new dataframe consisting of selected variables from the original HINTS dataset

hints_clndata <- subset(hints_data, select = c(HHID, PersonID, Age, AvgDrinksPerWeek, WeeklyMinutesModerateExercise, BMI, QualityCare, BirthGender, smokeStat, RaceEthn5))
```



 ##4. Univariate Visualization

### 4.I QualityCare and RaceEthn5
#Celia Lopez

```{r}
# Bar plot for QualityCare
qc <- table(hints$QualityCare)

barplot(qc,
        main = "Distribution of QualityCare",
        xlab = "QualityCare",
        ylab = "Count")

# Pie chart for RaceEthn5
race <- table(hints$RaceEthn5)

pie(race,
    main = "Distribution of RaceEthn5")
```

Interpretation: The bar plot shows the distribution of QualityCare categories. Some categories appear more frequently, indicating that many respondents reported similar levels of healthcare quality. The pie chart shows the proportion of individuals across RaceEthn5 categories.

### 4.II Age and BMI

```{r}
# Histogram for Age
hist(hints$Age,
     main = "Distribution of Age",
     xlab = "Age",
     ylab = "Frequency")

# Boxplot for BMI
boxplot(hints$BMI,
        main = "Distribution of BMI",
        ylab = "BMI")
```

Interpretation: The histogram shows the distribution of Age values among participants. Most individuals fall within a middle age range. The boxplot shows the spread of BMI values, with the median near the center and possible outliers.
